---
title: "Plasmid Alpha and Beta Diversity Using de novo Binary Data"
author: "Ilhan Cem Duru"
date: "2024-08-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, include=FALSE}
library(knitr)

library(BiocParallel)

library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(tidylog)
library(reshape2)
library(purrr)
library(magrittr)
library(kableExtra)
library(gridExtra)
library(cowplot)
library(patchwork)
library(GGally)
library(corrr)
library(energy)
library(MatrixCorrelation)
library(qqplotr)
library(ggridges)
library(coin)
library(dabestr)
library(phyloseq)
library(vegan)
library(DESeq2)
library(fido)
library(microbiome)
```

# Load and prepare data for phyloseq
## Create counts matrix
```{r, warning=FALSE}
# Load counts
plasmid_binary <- read.csv("all_results_cluster_simp_pivot.tsv", header = TRUE,
                       skipNul = TRUE, sep = "\t", as.is = TRUE)
#plasmid_binary
#str(plasmid_binary)
#row.names(plasmid_binary)
#class(plasmid_binary)
colnames(plasmid_binary)
#head(colnames(plasmid_binary))
#dim(plasmid_binary)
#plasmid_binary$phage

## Put this in `matrix` class for `phyloseq`'s `otu table`, and create it:
counts <- as.matrix(sapply(plasmid_binary, as.numeric))
#head(counts)
#head(colnames(counts))
rownames(counts) <- plasmid_binary$Plasmid
#head(counts)
#head(colnames(counts))

# delete/remove the newly created "phage" column:
counts <- counts[ , -grep("Plasmid", colnames(counts))]
#head(counts)
#head(colnames(counts))
#dim(counts)

otu.table <- otu_table(counts, taxa_are_rows = TRUE)
head(otu.table)
head(colnames(otu.table))
dim(otu.table)

#otu.table

```

## Create dummy tax table for TAX
```{r, warning=FALSE}
taxmat = matrix(sample(letters, 29096, replace = TRUE),
                nrow = nrow(otu.table), ncol = 7)
rownames(taxmat) <- rownames(otu.table)
colnames(taxmat) <- c("Domain", "Phylum", "Class", "Order",
                      "Family", "Genus", "Species")
#taxmat
TAX = tax_table(taxmat)
```

## Import the metadata:
```{r, warning=FALSE}
sampledata <- as.data.frame(read.csv
                            (file = "pd_meta_with_ffq_and_scfa_only_oursamples_3variables_ordered.csv",
                            header = TRUE, sep = ",", row.names = 1))
#rownames(sampledata)
#colnames(otu.table)
identical(rownames(sampledata), colnames(otu.table))
sampledata = sample_data(sampledata)
```

## Create physeq
```{r, warning=FALSE}
physeqfinal <- phyloseq(otu.table,TAX, sampledata)
#physeqfinal
#sample_data(physeqfinal)
#summary(sample_data(physeqfinal))

physeqfinal.2 <- subset_taxa(physeqfinal, taxa_sums(physeqfinal) >0)
#head(sort(taxa_sums(physeqfinal.2), decreasing = FALSE))

physeqfinal.2
physeqfinal

summary(sample_data(physeqfinal.2))
```

# Alpha Diversity for Group (C vs P)

```{r, warning=FALSE}
richness.table <- estimate_richness(physeqfinal.2, split = TRUE,
                                    measures = c("Observed", "Chao1",
                                    "ACE", "Shannon", "InvSimpson"))

head(richness.table)

richness.table$Group <- sample_data(physeqfinal.2)$Group
```

## Plot Alpha Diversity

```{r, warning=FALSE}
theme_set(theme_bw())

grid.arrange(
  ggplot(richness.table, aes(x = Group, y = Observed)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = Group, y = Chao1)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = Group, y = ACE)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = Group, y = Shannon)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = Group, y = InvSimpson)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  nrow = 1)
```

## Observed Richness Wilcoxon rank sum test (Group)
```{r, warning=FALSE}
wilcox.test(richness.table$Observed ~ sample_data(physeqfinal.2)$Group,
            conf.level = 0.95, conf.int = TRUE)
```

## CHAO1 Wilcoxon rank sum test (Group)
```{r, warning=FALSE}
wilcox.test(richness.table$Chao1 ~ sample_data(physeqfinal.2)$Group,
          conf.level = 0.95, conf.int = TRUE)
```
## ACE Wilcoxon rank sum test (Group)
```{r, warning=FALSE}
wilcox.test(richness.table$ACE ~ sample_data(physeqfinal.2)$Group,
          conf.level = 0.95, conf.int = TRUE)
```
## InvSimpson Wilcoxon rank sum test (Group)
```{r, warning=FALSE}
wilcox.test(richness.table$InvSimpson ~ sample_data(physeqfinal.2)$Group,
            conf.level = 0.95, conf.int = TRUE)
```
## Shannon Wilcoxon rank sum test (Group)
```{r, warning=FALSE}
wilcox.test(richness.table$Shannon ~ sample_data(physeqfinal.2)$Group,
            conf.level = 0.95, conf.int = TRUE)
```


# Alpha Diversity for Gender (F vs M)
```{r, warning=FALSE}
richness.table$gender <- sample_data(physeqfinal.2)$gender
```
## Plot Alpha Diversity for Gender (F vs M)
```{r, warning=FALSE}
theme_set(theme_bw())

grid.arrange(
  ggplot(richness.table, aes(x = gender, y = Observed)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = gender, y = Chao1)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = gender, y = ACE)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = gender, y = Shannon)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  ggplot(richness.table, aes(x = gender, y = InvSimpson)) +
    geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2, alpha =.5) +
    stat_summary(fun = mean, color = "darkred", geom = "point", shape = 18, size = 4),
  nrow = 1)
```

## Observed Richness Wilcoxon rank sum test (gender)
```{r, warning=FALSE}
wilcox.test(richness.table$Observed ~ sample_data(physeqfinal.2)$gender,
            conf.level = 0.95, conf.int = TRUE)
```

## CHAO1 Wilcoxon rank sum test (gender)
```{r, warning=FALSE}
wilcox.test(richness.table$Chao1 ~ sample_data(physeqfinal.2)$gender,
          conf.level = 0.95, conf.int = TRUE)
```
## ACE Wilcoxon rank sum test (gender)
```{r, warning=FALSE}
wilcox.test(richness.table$ACE ~ sample_data(physeqfinal.2)$gender,
          conf.level = 0.95, conf.int = TRUE)
```
## InvSimpson Wilcoxon rank sum test (gender)
```{r, warning=FALSE}
wilcox.test(richness.table$InvSimpson ~ sample_data(physeqfinal.2)$gender,
            conf.level = 0.95, conf.int = TRUE)
```
## Shannon Wilcoxon rank sum test (gender)
```{r, warning=FALSE}
wilcox.test(richness.table$Shannon ~ sample_data(physeqfinal.2)$gender,
            conf.level = 0.95, conf.int = TRUE)
```

# Alpha Diversity for BMI (Continuous variable)
```{r, warning=FALSE}
richness.table$BMI <- sample_data(physeqfinal.2)$BMI
```
## Plot Alpha Diversity for BMI (Continuous variable)
```{r, warning=FALSE}
common_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1),
)

grid.arrange(
  ggplot(richness.table, aes(x = BMI, y = Observed)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "BMI", y = "Observed") +
    scale_x_continuous(breaks = seq(20, 35, by = 5)) +  # Adjust range as needed
    common_theme,
  
  ggplot(richness.table, aes(x = BMI, y = Chao1)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "BMI", y = "Chao1") +
    scale_x_continuous(breaks = seq(20, 35, by = 5)) +
    common_theme,
  
  ggplot(richness.table, aes(x = BMI, y = ACE)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "BMI", y = "ACE") +
    scale_x_continuous(breaks = seq(20, 35, by = 5)) +
    common_theme,
  
  ggplot(richness.table, aes(x = BMI, y = Shannon)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "BMI", y = "Shannon") +
    scale_x_continuous(breaks = seq(20, 35, by = 5)) +
    common_theme,
  
  ggplot(richness.table, aes(x = BMI, y = InvSimpson)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "BMI", y = "InvSimpson") +
    scale_x_continuous(breaks = seq(20, 35, by = 5)) +
    common_theme,
  
  nrow = 1
)
```

## Observed Richness Spearman rank sum test (BMI)
```{r, warning=FALSE}
cor.test(richness.table$Observed, sample_data(physeqfinal.2)$BMI, method = "spearman")
```
## CHAO1 Richness Spearman rank sum test (BMI)
```{r, warning=FALSE}
cor.test(richness.table$Chao1, sample_data(physeqfinal.2)$BMI, method = "spearman")
```
## ACE Richness Spearman rank sum test (BMI)
```{r, warning=FALSE}
cor.test(richness.table$ACE, sample_data(physeqfinal.2)$BMI, method = "spearman")
```
## InvSimpson Richness Spearman rank sum test (BMI)
```{r, warning=FALSE}
cor.test(richness.table$InvSimpson, sample_data(physeqfinal.2)$BMI, method = "spearman")
```
## Shannon Richness Spearman rank sum test (BMI)
```{r, warning=FALSE}
cor.test(richness.table$Shannon, sample_data(physeqfinal.2)$BMI, method = "spearman")
```

# Alpha Diversity for Age (Continuous variable)
```{r, warning=FALSE}
richness.table$age <- sample_data(physeqfinal.2)$age_at_stool_collection
```
## Plot Alpha Diversity for Age (Continuous variable)
```{r, warning=FALSE}
common_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1),
)

grid.arrange(
  ggplot(richness.table, aes(x = age, y = Observed)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "age", y = "Observed") +
    scale_x_continuous(breaks = seq(50, 80, by = 10)) +  # Adjust range as needed
    common_theme,
  
  ggplot(richness.table, aes(x = age, y = Chao1)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "age", y = "Chao1") +
    scale_x_continuous(breaks = seq(50, 80, by = 10)) +
    common_theme,
  
  ggplot(richness.table, aes(x = age, y = ACE)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "age", y = "ACE") +
    scale_x_continuous(breaks = seq(50, 80, by = 10)) +
    common_theme,
  
  ggplot(richness.table, aes(x = age, y = Shannon)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "age", y = "Shannon") +
    scale_x_continuous(breaks = seq(50, 80, by = 10)) +
    common_theme,
  
  ggplot(richness.table, aes(x = age, y = InvSimpson)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x = "age", y = "InvSimpson") +
    scale_x_continuous(breaks = seq(50, 80, by = 10)) +
    common_theme,
  
  nrow = 1
)
```

## Observed Richness Spearman rank sum test (age)
```{r, warning=FALSE}
cor.test(richness.table$Observed,
         sample_data(physeqfinal.2)$age_at_stool_collection, method = "spearman")
```
## CHAO1 Richness Spearman rank sum test (age)
```{r, warning=FALSE}
cor.test(richness.table$Chao1,
         sample_data(physeqfinal.2)$age_at_stool_collection, method = "spearman")
```
## ACE Richness Spearman rank sum test (age)
```{r, warning=FALSE}
cor.test(richness.table$ACE,
         sample_data(physeqfinal.2)$age_at_stool_collection, method = "spearman")
```
## InvSimpson Richness Spearman rank sum test (age)
```{r, warning=FALSE}
cor.test(richness.table$InvSimpson,
         sample_data(physeqfinal.2)$age_at_stool_collection, method = "spearman")
```
## Shannon Richness Spearman rank sum test (age)
```{r, warning=FALSE}
cor.test(richness.table$Shannon,
         sample_data(physeqfinal.2)$age_at_stool_collection, method = "spearman")
```

# Distance-based multivariate analysis / beta diversity
## Jaccard distance calculation (Binary Data)
```{r, warning=FALSE}
#physeqfinal.2.clr <- microbiome::transform(physeqfinal.2, transform = "clr", target = "OTU",
#                                          shift = 0, scale = 1)

#do not apply clr transformation to binary data
physeqfinal.2.clr <- physeqfinal.2
physeqfinal.2.clr.jaccard_dist <- vegdist(t(as.data.frame(as.matrix(otu_table(physeqfinal.2.clr)))),
                                      method = "jaccard", binary = TRUE)

sample_data(physeqfinal.2.clr)$Group <- factor(sample_data(physeqfinal.2.clr)$Group,
                                              levels=c("C" , "P"),
                                              ordered = FALSE)
sample_data(physeqfinal.2.clr)$gender <- factor(sample_data(physeqfinal.2.clr)$gender,
                                              levels=c("F" , "M"),
                                              ordered = FALSE)
```

## Run the statistics for group variable
```{r, warning=FALSE}
adonis.Res.clr <- adonis2(physeqfinal.2.clr.jaccard_dist ~ 
                          sample_data(physeqfinal.2.clr)$Group,
                          perm = 10000,
                          na.action = na.exclude,
                          parallel = 10)

adonis.Res.clr
```

## Principal Coordinates Analysis (PCoA) for Group
```{r, warning=FALSE}
ord_clr <- ordinate(physeqfinal.2.clr, method = "PCoA",
                    distance = "jaccard", binary = TRUE, autotransform = FALSE,
                    trymax = 1000, maxit = 10000, sratmax = 0.999999,
                    previous.best, parallel = 10)

plot_ordination(physeqfinal.2.clr, ord_clr,
                type = "Samples", color = "Group") + geom_point(size = 3) +
  stat_ellipse(level = 0.95)
```

## Run the statistics for gender variable
```{r, warning=FALSE}
adonis.Res.clr <- adonis2(physeqfinal.2.clr.jaccard_dist ~ 
                          sample_data(physeqfinal.2.clr)$gender,
                          perm = 10000,
                          na.action = na.exclude,
                          parallel = 10)
adonis.Res.clr
```

## Principal Coordinates Analysis (PCoA) for Gender
```{r, warning=FALSE}
ord_clr <- ordinate(physeqfinal.2.clr, method = "PCoA",
                    distance = "jaccard", binary = TRUE, autotransform = FALSE,
                    trymax = 1000, maxit = 10000, sratmax = 0.999999,
                    previous.best, parallel = 10)
plot_ordination(physeqfinal.2.clr, ord_clr,
                type = "Samples", color = "gender") + geom_point(size = 3) +
  stat_ellipse(level = 0.95)
```

## Run the statistics for BMI variable
```{r, warning=FALSE}
adonis.Res.clr <- adonis2(physeqfinal.2.clr.jaccard_dist ~ 
                          sample_data(physeqfinal.2.clr)$BMI,
                          perm = 10000,
                          na.action = na.exclude,
                          parallel = 10)

adonis.Res.clr
```

## Principal Coordinates Analysis (PCoA) for BMI
```{r, warning=FALSE}
ord_clr <- ordinate(physeqfinal.2.clr, method = "PCoA",
                    distance = "jaccard", binary = TRUE, autotransform = FALSE,
                    trymax = 1000, maxit = 10000, sratmax = 0.999999,
                    previous.best, parallel = 10)
plot_ordination(physeqfinal.2.clr, ord_clr,
                type = "Samples", color = "BMI") + geom_point(size = 3) +
  stat_ellipse(level = 0.95)
```

## Run the statistics for Age variable
```{r, warning=FALSE}
adonis.Res.clr <- adonis2(physeqfinal.2.clr.jaccard_dist ~ 
                          sample_data(physeqfinal.2.clr)$age_at_stool_collection,
                          perm = 10000,
                          na.action = na.exclude,
                          parallel = 10)

adonis.Res.clr
```

## Principal Coordinates Analysis (PCoA) for Age
```{r, warning=FALSE}
ord_clr <- ordinate(physeqfinal.2.clr, method = "PCoA",
                    distance = "jaccard", binary = TRUE, autotransform = FALSE,
                    trymax = 1000, maxit = 10000, sratmax = 0.999999,
                    previous.best, parallel = 10)
plot_ordination(physeqfinal.2.clr, ord_clr,
                type = "Samples", color = "age_at_stool_collection") + geom_point(size = 3) +
  stat_ellipse(level = 0.95)
```

# Session Info for Reproducibility
```{r, warning=FALSE}
sessionInfo()
```